---
output: html_document
---
 
Health and Economic Impact of Weather Events in the US
======================================================
    
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

Data
====
The analysis was performed on
[Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp), provided by
[National Climatic Data Center](http://www.ncdc.noaa.gov/). 

The Storm data is from a comma-separated-value file available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)[47Mb].

There is also some documentation of the data available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

Unique Event Types as defined by NOAA can be found [here](http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf). Please Refer Section 7.


Synopsis
========

The analysis on the storm event database revealed that tornadoes are the most
dangerous weather event to the population health. The second most dangerous
event type is the excessive heat. The economic impact of weather events was
also analyzed. Flash floods and thunderstorm winds caused billions of dollars
in property damages between 1950 and 2011. The largest crop damage caused by
drought, followed by flood and hails.


Data Processing
===============
 
####Download Data

1. Downloaded Storm data file in local directory
2. I downloaded Unique Event Types from NOAA website and saved them into an excel file in local directory and named file event_type.csv.

At this point, i assume you have 2 files in your local directory.
If not, please download files before proceeding.
 
The first step is to read the data into a data frame.

```{r cache=TRUE}
#file_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
filename <- "repdata_data_StormData.csv.bz2"
#download.file(url=file_url, destfile=filename)
#downloadOn <- format(Sys.time(), "%b %d, %Y %X")
data <- read.csv(bzfile(filename))
```


####Pr-Processing of event type data
 
```{r}
library(dplyr)

# prepare event data
event_type_filename="event_type.csv"

event_type <- read.csv(event_type_filename, header=TRUE )
event_type <- mutate_each(event_type, funs(toupper))
names(event_type) <- c( "event", "desig")
# replace anything after \ with blank 
event_type <- as.data.frame(apply(event_type,2,function(x)gsub('/.*', '',x)))
# replace anything after ( with blank 
event_type <- as.data.frame(apply(event_type,2,function(x)gsub('\\(.*', '',x)))
#sub - with _ 
event_type <- as.data.frame(apply(event_type,2,function(x)gsub('-', '_',x)))
#remove leading blank
event_type <- as.data.frame(apply(event_type,2,function(x)gsub('^ ', '',x)))
#remove trailing blank
event_type <- as.data.frame(apply(event_type,2,function(x)gsub(' $', '',x)))
#sub blank with _ 
event_type <- as.data.frame(apply(event_type,2,function(x)gsub(' ', '_',x)))

# recode
event_type$ne <- event_type$event
event_type$ne <- gsub( "ASTRO.*", "ASTRONOMICAL", event_type$ne)

paste(" Number of Unique Storm Events as defined by NOAA :", length(unique(event_type$ne)))
```


####Pr-Processing of storm data

```{r}
# number of unique event types
paste("Count of Unique Storm Events in Storm Data :", length(unique(data$EVTYPE)))

#prepare storm data 
storm <- select( data,EVTYPE, FATALITIES, INJURIES,PROPDMG,CROPDMG,PROPDMGEXP,CROPDMGEXP )
storm <- mutate(storm, event=toupper(EVTYPE))
# replace anything after \ with blank 
storm$event <- gsub('/.*', '',storm$event)
# replace anything after ( with blank 
storm$event<- gsub('\\(.*', '',storm$event)
#sub - with _ 
storm$event <-  gsub('-', '_',storm$event)
#remove leading blank
storm$event<-  gsub('^ ', '',storm$event) 
#remove trailing blank
storm$event <-  gsub(' $', '',storm$event) 
#sub blank with _ 
storm$event<-  gsub(' ', '_',storm$event) 
 
# do not copy data that are difficult to map
storm <- storm[ -which(storm$event=="?"), ]
storm <- storm[ -which(grepl("SUMM|OTHER|NONE|RED|REM|WEATHER", storm$event)), ]
storm <- storm[ -which(grepl("EXCESSIVE|APACHE|SOUTH|NORTHERN|MILD", storm$event)), ]

 # recode/map data 
storm$event <- gsub( "WND|TSTM|GUST|WIND|FUNNEL", "HIGH_WIND", storm$event)
storm$event <- gsub( "URB|FLASH", "FLASH_FLOOD", storm$event)
storm$event <- gsub( "COAST|EROS|BEACH|DROW", "COASTAL_FLOOD", storm$event)
storm$event <- gsub( "DAM|STEAM|SEAS|WAVE|STREAM", "FLOOD", storm$event)
storm$event <- gsub( "WET|MICRO|CLOUD|BURST|WATER|WET|SHOWER", "HEAVY_RAIN", storm$event)
storm$event <- gsub( "LIGHT|PRECI|GLAZE|RAIN", "HEAVY_RAIN", storm$event)
storm$event <- gsub( "DRY|WARM|HOT|TEMP|RECORD|DRIEST|UNUSUAL", "EXCESSIVE_HEAT", storm$event)
storm$event <- gsub( "WILD|FIRE", "WILDFIRE", storm$event)
storm$event <- gsub( "HYPERTHERMIA|HYPO|EXTRE|MIX|FREEZ", "EXTREME_COLD", storm$event)
storm$event <- gsub( "LAND|MUD|SLIDE", "DEBRIS_FLOW", storm$event)
storm$event <- gsub( "ICE|ICY", "ICE_STORM", storm$event)
storm$event <- gsub( "TYP|HEAVY", "HURRICANE", storm$event)
storm$event <- gsub( "WINTRY", "WINTER_STORM", storm$event)
storm$event <- gsub( "HIGH", "TROPICAL_STORM", storm$event)
storm$event <- gsub( "BLOW|SAHARA|DUST", "BLIZZARD", storm$event)
storm$event <- gsub( "THUNDER|STORM", "THUNDERSTORM_WIND", storm$event)
storm$event <- gsub( "VOG|FOG", "DENSE_FOG", storm$event)
storm$event <- gsub( "SMOKE", "DENSE_SMOKE", storm$event)
storm$event <- gsub( "COOL|LOW", "COLD", storm$event)
storm$event <- gsub( "MARINE|SURF", "MARINE_HIGH_WIND", storm$event)
storm$event <- gsub( "SNOW", "HEAVY_SNOW", storm$event)
storm$event <- gsub( "TORNDAO", "TORNADO", storm$event)
storm$event <- gsub( "AVALANCE", "AVALANCHE", storm$event) 
storm$event <- gsub( "WAYTERSPOUT", "FLOOD", storm$event)
storm$event <- gsub( "LIGNTNING", "LIGHTNING", storm$event)
storm$event <- gsub( "VOLCANIC", "VOLCANIC_ASH", storm$event)

# sort the data
storm<-arrange(storm, event)

# add new event to be mapped with event_type
storm$stormClass <-   "" 

# match 
for( e in as.vector(unique(event_type$ne) )) { 
    storm$stormClass[grepl(e, storm$event)] <- e
}

# number of unique event typespost transformation of storm data
paste("Count of Unique Storm Events in Storm Data :", length(unique(storm$stormClass)))
```

 
After the cleaning, as expected, the number of unique event types reduce
significantly. For further analysis, the cleaned event types are used.


Most Harmful Events Data 
========================= 
    
   prepare FATALITIES, INJURIES,PROPDMG,CROPDMG Data
   
   To analyze economic damage:
   
   Property damage is represented with two fields, `PROPDMG` in dollars and the exponent `PROPDMGEXP`. 
   
   Crop damage is represented using two fields, `CROPDMG` and `CROPDMGEXP`.
   
   
```{r}
# prepare FATALITIES, INJURIES,PROPDMG,CROPDMG
storm$FATALITIES <- as.numeric( storm$FATALITIES )
storm$INJURIES <- as.numeric( storm$INJURIES )
```


####Prepare Property Damage Data


```{r}
# prepare  PROPDMG 
storm$PROPDMG <- as.numeric( storm$PROPDMG )

#clean prop  data 
storm$PROPDMGEXP<-  gsub('^ $', '',storm$PROPDMGEXP) 
storm$PROPDMGEXP <- gsub('[-?+]', 0, storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub('^$', 0, storm$PROPDMGEXP)

# sub letter with numbers
# h ->hundred, k - 1000, m -> million, b is billion...convert to million
storm$PROPDMGEXP <- gsub('[Hh]', .0001, storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub('[Kk]', .001, storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub('[Mm]', 1, storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub('[Bb]', 1000, storm$PROPDMGEXP)
 
#calc prop  damage in millions 
storm$PROPDMG <- as.numeric(storm$PROPDMG*as.numeric(storm$PROPDMGEXP))
```


####Prepare Crop Damage Data


```{r}
# prepare  CROPDMG
storm$CROPDMG <- as.numeric( storm$CROPDMG )

#clean prop and crop data 
storm$CROPDMGEXP<-  gsub('^ $', '',storm$CROPDMGEXP) 
storm$CROPDMGEXP <- gsub('[-?+]', 0, storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub('^$', 0, storm$CROPDMGEXP)

# sub letter with numbers
# h ->hundred, k - 1000, m -> million, b is billion...convert to million
storm$CROPDMGEXP <- gsub('[Hh]', .0001, storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub('[Kk]', .001, storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub('[Mm]', 1, storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub('[Bb]', 1000, storm$CROPDMGEXP)
 
#calc crop damage in millions 
storm$CROPDMG <- as.numeric(storm$CROPDMG*as.numeric(storm$CROPDMGEXP))
```


####Top 10 events that caused largest number of deaths are


```{r}
library(ggplot2)

# calc fatal data 
fatal  <- aggregate(  storm$FATALITIES , list( event =storm$stormClass), sum )
fatal <-fatal[order(fatal$x, decreasing = TRUE), ]
ggplot(fatal[1:10,], aes(x = reorder(event, x), y = x)) + 
    geom_bar(stat = "identity", fill = "blue", las = 3) +
    coord_flip() + 
    xlab("EVENT TYPE") + ylab("FATALITIES") +
    ggtitle("Fatalities by Top 10 Weather Events")
```


####Top 10 events that caused most number of injuries are


```{r}
# calc injury data 
injury  <- aggregate(  storm$INJURIES , list( event =storm$stormClass), sum )
colnames(injury) <- c("event", "total")
injury <-injury[order(injury$total, decreasing = TRUE), ]
ggplot(injury[1:10,], aes(x = reorder(event, total), y = total)) + 
    geom_bar(stat = "identity", fill = "red" ) +
    coord_flip() + 
    xlab("EVENT TYPE") + ylab("INJURIES") +
    ggtitle("Injuries by Top 10 Weather Events")
```

  
####Top 10 events that caused most property damage (in dollars) are  

 
```{r}
prop <- aggregate( storm$PROPDMG , list( event =storm$stormClass), sum )
names(prop) <- c("event", "loss")
prop <-prop[order(prop$loss, decreasing = TRUE), ]

ggplot(prop[1:10,], aes(x = reorder(event, loss), y = loss)) + 
    geom_bar(stat = "identity", fill = "green" ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("EVENT TYPE") + ylab("Property Loss( in Millions)") +
    ggtitle("Property Loss by Top 10 Weather Events")
```
 
 
####Top 10 events that caused most crop damage (in dollars) are as follows


```{r}
crop <- aggregate( storm$CROPDMG , list( event =storm$stormClass), sum )
names(crop) <- c("event", "loss")
crop <-crop[order(crop, decreasing = TRUE), ]

ggplot(crop[1:10,], aes(x = reorder(event, loss), y = loss)) + 
    geom_bar(stat = "identity", fill = "black" ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("EVENT TYPE") + ylab("Crop Loss( in Millions)") +
    ggtitle("Crop Loss by Top 10 Weather Events")
```
 

Results
=======

####Population Death & Injury

Based on NOAA data, Tornadoes cause most number of deaths and injuries among all event types. There are more than 5,000 deaths and more than 75,000 injuries in the last 60 years
in US, due to tornadoes. 
The other event types that are most dangerous with respect to population health
are wind and floods.

####Economic consequences to Property 

The most severe weather event in terms of property damage is the Flood. In the last
half century, the flood has caused more than 150 billion dollars damage. Other
severe property-damage-causing event types are hurricane and wind.
 
####Economic consequences to Crop

 The most severe weather event in terms of crop damage is the thunderstorm. In the last
half century, the it  has caused more than 8 billion dollars damage. Other
severe crop-damage-causing event types are tornado and fire.




 